A mixture of hidden Markov models to predict the lymphatic spread in head and neck cancer depending on primary tumor location

### 2024-10-11 manuscript.qmd
finish going through mixture & EM formalism

$$
P \left( \mathbf{D} \mid \boldsymbol{\theta}, \boldsymbol{\pi}\right) = \prod_k^K \left[P \left( \mathbf{X}=\boldsymbol{\xi}_k \mid \boldsymbol{\theta}, T_k \right) \right]
$$ {#eq-likelihood-HMM}

@@

If we could record the componet $m$ from which a patient $k$ was drawn from, we could store this information in a binary latent vector $\boldsymbol{\epsilon}_k$. The vector $\boldsymbol{\epsilon}_k$ has length $M$, with exactly the $m$-th element set to 1, indicating which model generated the patient’s data, and all other elements set to 0. Thus, for patient $k$, the latent variable $\boldsymbol{\epsilon}_k$ can be interpreted as a categorical indicator variable that encodes the assignment to one of the $M$ lymphatic spread models. Unfortunately, the latent variables are hidden (unobserved) and must be inferred.

The joint probability of the observed data $\mathbf{D}$ (i.e., patient data) and the latent variables $\boldsymbol{\epsilon}$ is given by:

$$
P(\mathbf{D}, \boldsymbol{\epsilon} | \boldsymbol{\theta}, \boldsymbol{\pi}) = \prod_{k}^{K} \prod_{m}^{M} \left[ \pi_m^{s_k} P(\mathbf{D}_k | \boldsymbol{\theta}_m, \boldsymbol{\pi}) \right]^{\epsilon_{k}^m}
$${#eq-complete-likelihood}

@@

* $\pi_m^{s_k}$ is the mixing parameter for patient $k$ with a primary tumor in subsite $s_k$ and model $m$,
* $P(\mathbf{D}_k | \boldsymbol{\theta}_m)$ is the likelihood of patient $k$'s data given that it was generated by model $m$,
* $\boldsymbol{\theta}_m$ represents the parameters of model $m$,
* $\epsilon_{k}^m$ is the latent variable that indicates whether patient $k$ was generated by model $m$.

**Incomplete data likelihood which can probably be skipped**

Since the latent variables $\boldsymbol{\epsilon}_k$ are unobserved, the incomplete complete data likelihood marginalizes over all possible latent assignments to reflect the uncertainty about which model generated each patient’s data:

$$
P(\mathbf{D} | \boldsymbol{\theta}, \boldsymbol{\pi}) = \sum_{\boldsymbol{\epsilon}} P(\mathbf{D}, \boldsymbol{\epsilon} | \boldsymbol{\theta}, \boldsymbol{\pi})
$$

This allows us to integrate out the latent variables $\boldsymbol{\epsilon}$, as their true values are unknown.

**Continues from here**

The latent variables $\boldsymbol{\epsilon}_k$ are unknown and must be inferred from the observed data. The mixture model assumes that each patient is generated by exactly one of the $M$ models. To do so a probabilistic distribution over the latent variables, given the observed data and model parameters is computed, the posterior distribution:

$$
P(\boldsymbol{\epsilon}_k | \mathbf{D}_k, \boldsymbol{\theta}, \boldsymbol{\pi})
$$

describes the probability that patient $k$ was generated by each of the $M$ models, allowing for soft assignments where each patient is associated with multiple models with certain probabilities, as opposed to being strictly assigned to just one. Given the probability distribution for the latent variables $\boldsymbol{\epsilon}_k$, we compute the expectation value $\mathbb{E}[\boldsymbol{\epsilon}_k] = \boldsymbol{\gamma}(\boldsymbol{\epsilon}_k)$ for each patient to then apply the complete data likelihood given in @eq-complete-likelihood.

The likelihood of the complete dataset is then expressed as:

$$
P \left( \mathbf{D} \mid \boldsymbol{\theta}, \boldsymbol{\pi}\right) = \prod_k^K \left[ \prod_{m}^M \pi_m^{s_k} P_m \left( \mathbf{X}=\boldsymbol{\xi}_k \mid \boldsymbol{\theta}_m, T_k \right) \right]^{\gamma(\epsilon_k^m)} 
$$ {#eq-complete-likelihood-explicite}

In @eq-complete-likelihood we compute the likelihood of observing the patient's state $\boldsymbol{\xi}_k$ with t-stage $T_k$ under model $m$ similarly to @eq-likelihood-HMM. However, we weight each model with the mixing parameter $\pi_m^{s_k}$. Since the mixing parameters are subsite dependent, we differ between each subsite.

It is important to mention, that the latent variables $\boldsymbol{\epsilon}$ are only used for parameter estimation. The subsite specific mixture parameters $\boldsymbol{\pi}^s$ are the only parameters which are used for later risk predictions. In @fig-model-simple the mixture parameters are illustrated. Subsites with different spread patterns, such as Gum (C03) and Base of tongue (C01), are expected to get different model assignments. Nonetheless, the latent variables for two patients with the same diagnosis, i.e. same involvement $\boldsymbol{\xi}$ and and t-stage $T$, but different subsites are the same.

@@

The model now contains two sets of parameters: the tumor spread probabilities $\mathbf{\theta_m}$ for each model, and the mixing coefficients $\pi_m^s$. The posterior distribution over these parameters $P(\boldsymbol{\theta},\boldsymbol{\pi}|\mathbf{D})$ can be estimated assuming a uniform prior on the interval [0,1] for all parameters. However, $P(\boldsymbol{\theta},\boldsymbol{\pi}|\mathbf{D})$ is a multimodal-distribution as one can permute the different models and achieve the same result. To address this problem, we apply an expectation-maximization (EM) algorithm.

## Expectation-Maximization (EM) Algorithm
To solve for the model parameters, we apply the EM algorithm, which iterates between two steps: the **Expectation (E) step** and **the Maximization (M) step**.

#### Expectation Step

In the E-step, we compute the expected value of the latent variables $\boldsymbol{\epsilon}$, which represent the probabilities of a patient $k$ originating from one of the models $m$. Given the current estimates of $\boldsymbol{\theta}$ and $\boldsymbol{\pi}$, the expectation value $\gamma(\epsilon_k^m)$ is calculated as:

$$
\mathbb{E}[\epsilon^m_k] = \gamma(\epsilon^m_{k}) = \frac{\pi_m^{s_k} P(\mathbf{X} = \boldsymbol{\xi}_k \mid \boldsymbol{\theta_m}, T_k)}{\sum_{n}^M \pi_n^{s_k} P(\mathbf{X} = \boldsymbol{\xi}_k \mid \boldsymbol{\theta_n}, T_k)}
$${#eq-expectation-step}

#### Maximization Step

In the M-step, the model parameters are updated by maximizing the complete data likelihood in equation @eq-complete-likelihood-explicite. For better handling, we are optimizing the log likelihood:

$$
\ln P(\mathbf{D}, \boldsymbol{\epsilon} \mid \boldsymbol{\theta}, \boldsymbol{\pi}) = \sum_k^K \sum_{m}^M \gamma(\epsilon^m_{k}) \left( \ln \pi_m^{s_k} + \ln P(\mathbf{X} = \boldsymbol{\xi}_k \mid \boldsymbol{\theta_m}, T_k) \right)
$${#eq-maximization-step}

In this step, the mixture parameters $\pi_m^s$ and the model parameters $\boldsymbol{\theta}_m$ can be optimized separately. The mixture parameters $\pi_m^s$ can be updated analytically:

$$
\pi_m^s = \frac{1}{N^s} \sum_k^{N^s} \gamma(\epsilon^m_k)
$${#eq-mixture-maximization}

On the other hand, the model parameters $\boldsymbol{\theta_m}$ are optimized component-wise using an optimization algorithm, as an analytic solution is not feasible. The updated log-likelihood for each model is:

$$
\ln P(\mathbf{D}, \boldsymbol{\epsilon} \mid \boldsymbol{\theta}_m) = \sum_{k}^K \gamma(\epsilon^m_{k}) \ln P(\mathbf{X} = \boldsymbol{\xi}_i \mid \theta_m, T_k)
$$

By iterating these steps, the EM algorithm converges to a (local) maximum of the complete data likelihood @eq-complete-likelihood.

>>>>

$$
P \left( \mathbf{D} \mid \boldsymbol{\theta} \right) = \prod_k^K P \left( \mathbf{X}=\boldsymbol{\xi}_k \mid \boldsymbol{\theta}, T_k \right)
$$ {#eq-likelihood-HMM}

@@

If we could record the componet $m$ from which a patient $k$ was drawn from, we could store this information in a binary latent vector $\boldsymbol{\epsilon}_k$. The vector $\boldsymbol{\epsilon}_k$ has length $M$, with exactly the $m$-th element set to 1, indicating which model generated the patient’s data, and all other elements set to 0. Thus, for patient $k$, the latent variable $\boldsymbol{\epsilon}_k$ can be interpreted as a categorical indicator variable that encodes the assignment to one of the $M$ lymphatic spread models. Typically in mixture models, this so-called _latent variable_ is unknown. However, it can be inferred and is useful for inferring the models' parameters.

The joint probability of the observed data $\mathbf{D}$ (i.e., patient data) and the latent variables $\boldsymbol{\epsilon}$ - sometimes called the _complete data likelihood_ - is given by:

$$
P(\mathbf{D}, \boldsymbol{\epsilon} \mid \boldsymbol{\theta}, \boldsymbol{\pi}) = \prod_{k}^{K} \prod_{m}^{M} \left[ \pi_m^{s_k} P(\mathbf{D}_k \mid \boldsymbol{\theta}_m) \right]^{\epsilon_{k}^m}
$$ {#eq-complete-likelihood}

@@

- $\pi_m^{s_k}$ is the mixing coefficient for subsite $s_k$ (where patient $k$ has their tumor) and model $m$,
- $P(\mathbf{D}_k \mid \boldsymbol{\theta}_m)$ is the likelihood of patient $k$'s diagnosis, given that it was generated by model $m$,
- $\boldsymbol{\theta}_m$ represents the parameters of model $m$,
- $\epsilon_{k}^m$ is the latent variable that indicates patient $k$ was generated by model $m$.

::: {.callout-note}

## TODO: Discuss if we need this
<!-- Not sure we can skip it. We want to maximize this w.r.t. the model parameters. -->

In contrast, the _incomplete data likelihood_ marginalizes over all possible latent assignments to reflect the uncertainty about which model generated each patient’s data:

$$
P(\mathbf{D}, \mid \boldsymbol{\theta}, \boldsymbol{\pi}) = \prod_{k}^{K} \sum_{m}^{M} \pi_m^{s_k} P(\mathbf{D}_k \mid \boldsymbol{\theta}_m)
$$ {#eq-incomplete-likelihood}

Ultimately, we want to find the parameters which maximize this likelihood function for the given data.

Note the summation inside the product. This structure of mixture models makes naive inference difficult, because the logarithm of this quantity is expensive to compute and not easy to differentiate. Thus, inferring the latent assignment is helpful, because the complete data (log-)likelihood (@eq-complete-likelihood) does not suffer from this shortcoming.

:::

To infer the latent assignment $\boldsymbol{\epsilon}_k$, we start with its distribution, given the observed data and model parameters:

$$
\gamma(\epsilon_k^m) := \mathbb{E}[\epsilon_k^m] = \frac{\pi_m^{s_k} P \left( \mathbf{D}_k \mid \boldsymbol{\theta}_m \right)}{\sum_{j \leq M} \pi_j^{s_k} P \left( \mathbf{D}_k \mid \boldsymbol{\theta}_j \right)}
$$ {#eq-responsibilities}

This expectation value - often called the _responsibility_ - describes the probability that patient $k$ was generated by model $m$. It can be used to compute the expected complete data likelihood:

$$
\mathbb{E}_{\boldsymbol{\epsilon}} \left[ P \left( \mathbf{D}, \boldsymbol{\epsilon} \mid \boldsymbol{\theta}, \boldsymbol{\pi} \right) \right] = \prod_{k=1}^K \prod_{m=1}^M \left[ \pi_m^{s_k} P \left( \mathbf{D}_k \mid \boldsymbol{\theta}_m \right) \right]^{\gamma(\epsilon_k^m)}
$$ {#eq-expected-complete-likelihood}

This expected complete data likelihood has the same tractable form as @eq-complete-likelihood and thus allows us to infer both the mixing coefficients $\boldsymbol{\pi}$ as well as each component model's parameters $\boldsymbol{\theta}_m$ using straightformward inference. Also, these two sets of parameters are the only ones used for the later risk prediction. The responsiblities $\gamma(\epsilon_k^m)$ are only used during inference.

In @fig-model-simple the mixture coefficients are illustrated. Subsites with different spread patterns, such as Gum (C03) and Base of tongue (C01), are expected to get different model assignments. Nonetheless, the latent variables for two patients with the same diagnosis, i.e. same involvement $\boldsymbol{\xi}$ and and t-stage $T$, but different subsites are the same.

@@

## Expectation-Maximization (EM) Algorithm

To actually find the mixing coefficients $\boldsymbol{\pi}$ and all models' parameters $\boldsymbol{\theta}_m$ that maximize @eq-incomplete-likelihood, we follow an iterative approach called _expectation-maximization_ or _EM-algorithm_. With arbitrarily initialized starting parameters, we alternate between the following two steps:

1. In the **E**xpectation step, we compute the responsibilities $\gamma(\epsilon_k^m)$, which represent the probabilities of a patient $k$ originating from one of the models $m$, given the current estimates of $\boldsymbol{\theta}$ and $\boldsymbol{\pi}$, as given in @eq-responsibilities.
2. During the **M**maximization step, we find a new set of parameters that maximize the new expected complete data likelihood (@eq-expected-complete-likelihood). For the mixture coefficients, we can even find an analytic solution to the new maximum:
   $$
   \pi_m^s = \frac{1}{|K_s|} \sum_{k \in K_s} \gamma(\epsilon_k^m)
   $$
   where we sum over the set of all patients $K_s$ with their tumor in subsite $s$.\
   The models' new parameters are found by numerically maximizing the respective likelihood, weighted by the responsibilities:
   $$
   \ln P(\mathbf{D}, \boldsymbol{\epsilon} \mid \boldsymbol{\theta}_m) = \sum_k^K \gamma(\epsilon^m_{k}) \big[ \ln \pi_m^{s_k} + \ln P \left( \mathbf{D}_k \mid \boldsymbol{\theta}_m \right) \big]
   $$ {#eq-maximization-step}

By iterating these steps, the EM algorithm is guaranteed to converge to a (local) maximum of the complete data likelihood (@eq-complete-likelihood).
